Interactive Model Serving for Regulatory Genomics

### 2016-12-17 paper.tex
figure formatting and grammer

\maketitle
\begin{abstract}
Current pipelines for training machine learning models on genomic and epigenetic datasets lack resources to scale to current datasets sizes. Furthermore, pipelines for testing and evaluating these machine learning models are segmented, as they require users to separately test pipelines and evaluate their methods using disjunct software. In this paper we survey current workflows for training, testing and evaluating genomic machine learning models and provide an alternative solution to connecting the dots between these processes. More specifically, present a complete fast and interactive pipeline that allows users to save prebuilt models and evaluate these models in a visual environment with low latency.
\end{abstract}

@@

\section{Introduction}
Regulatory genomics is a field of research that works to connect the function of DNA sequence, DNA interacting proteins and cell environment with gene expression. Understanding and predicting regulatory function is important, as it allows clinicians to predict the effect of an individual's genetic variants on their health and well being. Currently, there exist many different computational methods for prediction of regulatory function, including convolutional neural networks and SVM classification problems \cite{alipanahi2015predicting}, \cite{kelley2016basset}.  Datasets used to predict regulatory function are massive, reaching 15 TB in the ENCODE consortium \cite{encode2004encode}. However, these datasets are noisy, and thus analysis of regulatory predictions often require a human in the loop to visually interpret predictions against ground truth datasets produced in datasets such as ENCODE. \\

Current pipelines for training and analyzing regulatory genomic models include a segmented group of steps that are exclusive and require preprocessing between each stage. For example, the state-of-the-art pipeline involves executing a locally run algorithm, such as a kernel method or neural network, built for a single machine. Users who are interested in predicting on a particular sequence face multiple challenges. Because the state-of-the-art methods are not built to scale past a single machine, users must first sub-select a number of regions of the genome to train on specific to their hardware capabilities. After sub-selecting the data and training on that data, the user would then load these regions into a single node visualization tool, such as \cite{igv2011}, and repeat the preprocessing step to validate their model. A demonstration of the workflow pipeline required using currently available tools can be shown in Figure~\ref{fig:strawmanPipeline}.

@@

Currently, there is no infrastructure available for users to upload and predict on their sequences while receiving real time model serving. In practice, this infrastructure is not available, due in part to the difficulty in serving models at a large scale. Genomic data typically ranges on the order of 100's of MB to 10's of GB of data. The models are typically built with 10's of GB to 100's of GB of data, creating a unique challenge when trying to achieve interactive model serving speeds.

@@

Here, we provide an end-to-end pipeline that efficiently computes and predicts on regulatory genomics while maintaining seamless integration between processing steps. In our previous work, we use a scalable and parallel algorithm to efficiently compute protein binding sites on the whole genome, bypassing user defined subselection methods required for local algorithms \cite{tfbinding}. Here, we utilize this algorithm to interactively predict on new datasets, and provide a full analysis pipeline to analyze results. Our pipeline allows users to interactively predict on regions of the genome, removing file preprocessing steps required in current pipelines. We compare our pipeline to the current state of the art approaches intended for a local machine, and then analyze pipeline latency at scale on the whole genome.

@@

\subsection{Algorithms}
Current algorithms for predicting on regulatory genomic datasets include neural networks and SVM classification.
\begin{table}[t]
\centering
\caption{Featurization Techniques}
\label{table:featurization}
\begin{tabular}{| L{2cm} | L{1.45cm} | L{2cm} | L{1.75cm} |}
\hline
Featurization Technique & Datasets & Description & Usage
\\ \hline
Read Wavelets \newline
\cite{wavelets}         & DNA-seq, RNA-seq           & Iteratively quantifies read pileup                                 & Identification of genomic footprints \\ \hline
Kernel Approximation
\cite{tfbinding} & Sequence, DNA-seq,
RNA-seq & Approximates kernel matrix through random features & Generic, Widely applicable \cite{alipanahi2015predicting}                                       \\ \hline
K-mer counts & Sequence & Counts the number of times each motif
appears in a sequence of length k & Identification
of motifs, quantification of sequence similarity \\ \hline
\end{tabular}
\end{table}
These methods train on small portions of the genome. For example, a neural net implementation DeepBind sub-selects 10,000 sequences to predict protein binding sites. Similarly, Kernel SVM uses 10,000 sequence to learn a classifier. Both tools are intended to run on a single machine. \\

\subsection{Evaluation Tools}
Main interative tools for evaluating algorithms discussed above include visualization tools, such as University of Santa Cruz Genome Browser and IGV (\cite{ucscbrowser}, \cite{igv2011}). IGV is a desktop application that allows visualization of a wide variety of genomic datasets. IGV is used by clinicians and researchers to verify anomalies in their data and verify predictions to original raw genomic datasets. IGV does not support integrated visualization of predictive models, so users attempting to perform visual analysis of models using heterogeneous datasets must predict outside of the visualization tool.

\section{Architecture}
In this section, we outline our design for an end-to-end machine learning pipeline on epigenetic datasets. Our implementation builds upon previously built distributed genomics tools such as ADAM, Mango and Endive, which are described below. Finally, we will demonstrate how these tools can be connected together, with our new machine learning tool, to create an end-to-end pipeline for machine learning tasks. \\

@@

\subsection{Mango: Genome Visualization}
Mango is a distributed genome visualization browser that selectively materializes and organizes genomic data to provide fast and fine grained queries on datasets. Mango materializes data from persistent storage as the user requests different regions of the genome. This data is efficiently partitioned and organized in memory using a specialized RDD structure built upon interval trees. This interval based organizational structure supports ad hoc queries, enabling exploratory interaction with genomic data.  Mango is built on top of Spark and ADAM. A demonstration of Mango architecture can be seen in Figure X. Mango is implemented in three layers, namely, the cluster, server and client. In the cluster layer, Mango is built on ADAM and Spark, allowing users to perform and visualize fine grained queries on large genomic files. The server layer reduces and samples queries so that users can visualize their queries in an efficient and interpretable manner.

@@

@@

@@

\section{Experiments}
For preliminary results, we train a model predicting protein binding sites using the kernel approximation featurization method described in Table~\ref{table:featurization}  from ENCODE datasets using DNA sequence. This model was solved using Keystone ML’s Weighted Least Squares estimator, which accounts for large class imbalance of positive binding sites across the genome. This model was trained both locally on a small dataset as well as in a distributed environment on a large (200 million) dataset. These models generated a 4 MB model to be tested on new datasets. This model was loaded into the Mango genome browser and was used to interactively predict on DNA reference sequences as the user scrolled to new regions of the browser.

@@

Initially, our results were on the order of 6 seconds, which is sub-optimal for an application with a user interface. The results in Figure \# show a difference between our implementation with and without the binary pre-fetching approach. Binary pre-fetching allows us to rule out a significant part of the Servable Chunk.

>>>>

\maketitle
\begin{abstract}
Current pipelines for training machine learning models on genomic and epigenetic datasets lack resources to scale to current datasets sizes. Furthermore, pipelines for testing and evaluating these machine learning models are segmented, as they require users to separately test pipelines and evaluate their methods using disjunct software. In this paper we survey current workflows for training, testing and evaluating genomic machine learning models and provide an alternative solution to connecting the dots between these processes. More specifically, we present a complete fast and interactive pipeline that allows users to save prebuilt models and evaluate these models in a visual environment with low latency.
\end{abstract}

@@

\section{Introduction}
Regulatory genomics is a field of research that works to connect the function of DNA sequence, DNA interacting proteins and cell environment with gene expression. Understanding and predicting regulatory function is important, as it allows clinicians to predict the effect of an individual's genetic variants on their health and well being. Currently, there exist many different computational methods for prediction of regulatory function, including convolutional neural networks and kernels with SVM classification \cite{alipanahi2015predicting}, \cite{kelley2016basset}.  Datasets used to predict regulatory function are massive, reaching 15 TB in the ENCODE consortium \cite{encode2004encode}. However, these datasets are noisy, and thus analysis of regulatory predictions often require a human in the loop to visually interpret predictions against ground truth datasets produced in datasets such as ENCODE. \\

Current pipelines for training and analyzing regulatory genomic models include a segmented group of steps that are exclusive and require preprocessing between each stage. For example, the state-of-the-art pipeline involves executing a locally run algorithm, such as a kernel method or neural network, built for a single machine. Users who are interested in predicting on a particular sequence face multiple challenges. Because the state-of-the-art methods are not built to scale past a single machine, users must first subselect a number of regions of the genome to train on specific to their hardware capabilities. After subselecting and training on that data, the user would then load these regions into a single node visualization tool, such as \cite{igv2011}, and repeat the preprocessing step to validate their model. A demonstration of the workflow pipeline required using currently available tools can be shown in Figure~\ref{fig:strawmanPipeline}.

@@

Currently, there is no infrastructure available for users to upload and predict on their sequences in real time. In practice, this infrastructure is not available, due in part to the difficulty of serving models at a large scale. Genomic data typically ranges on the order of 100's of MB to 10's of GB of data. The models are typically built with 10's of GB to 100's of GB of data, creating a unique challenge when trying to achieve interactive model serving speeds.

@@

@@

\subsection{Algorithms}
Current algorithms for predicting on regulatory genomic datasets include neural networks and string kernels with SVM classification.

These methods train on small portions of the genome. For example, a neural net implementation DeepBind subselects 10,000 sequences to predict protein binding sites. Similarly, Kernel SVM uses 10,000 sequence to learn a classifier. Both tools are intended to run on a single machine. \\

\subsection{Evaluation Tools}
Main interative tools for evaluating algorithms discussed above include visualization tools, such as University of Santa Cruz Genome Browser and IGV \cite{ucscbrowser}, \cite{igv2011}. IGV is a desktop application that allows visualization of a wide variety of genomic datasets. IGV is used by clinicians and researchers to verify anomalies in their data and verify predictions on original raw genomic datasets. IGV does not support integrated visualization of predictive models, so users attempting to perform visual analysis of models using heterogeneous datasets must predict outside of the visualization tool.

\section{Architecture}
In this section, we outline our design for an end-to-end machine learning pipeline on epigenetic datasets. Our implementation builds upon previously built distributed genomics tools such as ADAM, Mango and Endive, which are described below. Finally, we will demonstrate how these tools can be connected together to create an end-to-end pipeline for machine learning tasks. \\

@@

In the use case of training a machine learning model on genomic data, ADAM is used to format training
 datasets into standardized schemas that can be saved and read by the hadoop ecosystem. These formats can then be read by ADAM and joined or aggregated to preprocess datasets into records read by a machine learning model.

\subsection{Mango: Genome Visualization}

@@

Mango is a distributed genome visualization browser that selectively materializes and organizes genomic data to provide fast and fine grained queries on datasets. Mango materializes data from persistent storage as the user requests different regions of the genome. This data is efficiently partitioned and organized in memory using a specialized RDD structure built upon interval trees. This interval based organizational structure supports ad hoc queries, enabling exploratory interaction with genomic data.  Mango is built on top of Spark and ADAM. A demonstration of Mango architecture can be seen in Figure~\ref{fig:mango}. Mango is implemented in three layers, namely, the cluster, server and client. In the cluster layer, Mango is built on ADAM and Spark, allowing users to perform and visualize fine grained queries on large genomic files. The server layer reduces and samples queries so that users can visualize their queries in an efficient and interpretable manner.

@@

\begin{table}[t]
\centering
\caption{Featurization Techniques}
\label{table:featurization}
\begin{tabular}{| L{2cm} | L{1.45cm} | L{2cm} | L{1.75cm} |}
\hline
Featurization Technique & Datasets & Description & Usage
\\ \hline
Read Wavelets \newline
\cite{wavelets}         & DNA-seq, RNA-seq           & Iteratively quantifies read pileup                                 & Identification of genomic footprints \\ \hline
Kernel Approximation
\cite{tfbinding} & Sequence, DNA-seq,
RNA-seq & Approximates kernel matrix through random features & Generic, Widely applicable \cite{alipanahi2015predicting}                                       \\ \hline
K-mer counts & Sequence & Counts the number of times each motif
appears in a sequence of length k & Identification
of motifs, quantification of sequence similarity \\ \hline
\end{tabular}
\end{table}

@@

\section{Experiments}
For preliminary results, we train a model predicting protein binding sites using the kernel approximation featurization method described in Table~\ref{table:featurization}  from ENCODE datasets using DNA sequence. This model was solved using Keystone ML's Weighted Least Squares estimator, which accounts for large class imbalance of positive binding sites across the genome. This model was trained both locally on a small dataset as well as in a distributed environment on a large (200 million) dataset. These models generated a 4 MB model to be tested on new datasets. This model was loaded into the Mango genome browser and was used to interactively predict on DNA reference sequences as the user scrolled to new regions of the browser.

@@

In the bwowser, we test on region sizes of 2000 bp, 2000 bp, 4000bp, 8000bp and 16000bp. Initially, our results were on the order of 6 seconds, which is sub-optimal for an application with a user interface. The results in Figure~\ref{fig:igvcomp} show a difference between our implementation with and without the binary pre-fetching approach. Binary pre-fetching allows us to rule out a significant part of the Servable Chunk.
